The core transcriptome of mammalian placentas and the divergence of expression with placental shape

### 2016-11-28 manuscript/manuscript.Rnw
the dog fetus died too
use the long names everywhere

Previous studies of some species (i.e., \textit{Homo sapiens},
\textit{Mus musculus}, \textit{Bos taurus}) have provided some
preliminary answers to these questions. They revealed a stunning
divergence in the genetics of the placenta, especially in later stages
of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
For example, each lineage has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine}.
These include independent expansions of the prolactins in \mouse,
\textit{Spalax carmeli}, \textit{Spalax galili}, and \cow, \acp{PSG}
in \human and \mouse, \acp{PAG} in \cow, genes related to the
thymus-specific serine protease \textit{PRSS16} in \textit{Loxodona
  africana} (in this work), and the expansion of the growth hormone
and chorionic gonadotropin beta genes in anthropoid primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between \human and
\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}.

To address our questions, we 
sequenced and quantified the total RNA of mammalian placentas at term,
including \human, \textit{Pan paniscus} (the sister
species to \textit{P. troglodytes}), \textit{Ateles fusciceps},
\mouse, \galili, \carmeli, \textit{Canis familiaris}, \cow, \elephant,
\textit{Dasypus novemcinctus}, and \textit{Monodelphis domesticus}.
These species represent four major branches of eutherian mammals
(Euarchontoglires, Laurasiatheria, Afrotheria, and Xenartha) as well
as marsupials. These placentas also represent a wide array of
placental morphologies as well as traits such as number and size of
offspring and length of gestation (\cref{tab:gestational_periods}). We
use this data, along with existing RNA-Seq data from \mouse
\cite{Wang.ea2011:surveyfornovelimprinted}, \human
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual},
\textit{Equus caballus}
\cite{Coleman.ea2010:Structuralannotationofequine}, \textit{Ovis
  aries} \cite{Archibald.ea2010:sheepgenomereferencesequence}, and
\textit{Sus scrofa} \cite{Ahanda.ea2012:Predictionofaltered3-} to: 1)
identify the non-housekeeping genes which are expressed in the
placentas of all species analyzed and therefore represent the
\emph{core functionality of the placenta}; 2) identify genes whose
expression level changes significantly in each of the morphology
types.

@@

Fetal tissue was collected from the placenta of nine mammalian
species (\human [human], \bonobo [bonobo], \smonkey [spider monkey], \dog [domestic dog], \cow
[domestic cow], \elephant [African savanna elephant], \opossum [gray
short-tailed opossum], \galili [blind mole rat from Upper Galilee],
\carmeli [blind mole rat from northern Israel]
\cite{Nevo2001:adaptiveradiationofblindsubterraneanmole}). Information
on specimen ID, and collector is shown for each sample in Supplemental
\cref{tab:srr_numbers}. In the case of \bonobo, \cow, and \elephant,
the placenta was collected at term upon expulsion from the mother
following delivery. Placental tissue from \human was collected upon
delivery via a cesarean section. Near-term placental tissue from \dog
was collected upon the death of the mother from a car accident.
Placental tissue from \opossum was sampled at fetal stage 32, roughly
one day prior to birth. Only the fetal portion of the placenta was
sampled at the maternal-fetal interface (i.e. villous portion for
\human, \bonobo, \dog, and \elephant; cotyledon for \cow; diffuse
placenta for \opossum). All tissues were collected and stored in
RNAlater (Life Technologies, Carlsbad, CA) until RNA isolation could
be performed. Tissues were homogenized in Trizol, and total RNA was
isolated following the manufacturer's instructions. RNA was further
purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in conjunction
with the RNase-Free DNase Set (Qiagen, Valencia, CA), following the
manufacturer’s protocol to remove any contaminating DNA from the
sample. Total RNA was then evaluated on a Nanodrop 1000 (Thermo
Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100 (Santa
Clara, CA) to confirm concentration and quality of the isolation. RNA
quality was assessed using A260/A280 and A260/A230 absorbance values.

@@

For more complete evolutionary comparisons additional sequences of
placenta in \mouse
(n=\Sexpr{star.logs[grepl("musculus",species),length(unique(file))]})
\cite{Wang.ea2011:surveyfornovelimprinted}, \human
(n=\Sexpr{star.logs[grepl("sapiens",species),length(unique(file))]})
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} \horse
\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \pig
\cite{Ahanda.ea2012:Predictionofaltered3-} were obtained from the
\ac{SRA} (Supplemental \cref{tab:srr_numbers}) and analyzed using the
same methodology as RNAs sequenced specifically for this study.

@@

\begin{sloppypar}
Reads quality was evaluated using FASTQC v0.11.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Transcriptomes without a suitable reference genome (\smonkey, \galili,
\carmeli; italic reference in Supplemental
\cref{tab:ref_genome_aligned_species}) were assembled using Trinity
v2014-07-17 \cite{Haas.ea2013:Denovotranscriptsequence} and annotated
using Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against
either the \human or \mouse transcriptome (Supplemental
\cref{tab:ref_genome_aligned_species}), and the
the match with the highest bitscore was chosen as the annotation.
Transcripts where the highest scoring match had a bitscore below 40
were left un-annotated.
% 
All transcriptomes were aligned to the reference (or
assembled) genome using STAR v2.4.2a~%
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} (Supplemental
\cref{fig:transcriptome_statistics,fig:transcriptome_reads_vs_read_length})
and expression values in \ac{FPKM}
% \cite{Trapnell.ea2010:Transcriptassemblyandquantification}
were
quantified using Cufflinks v2.2.1
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions} (Supplemental
\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}).
%
The genome of the closely-related \textit{P. troglodytes} (separation
$\approx 1$ million years
\cite{Prüfer.ea2012:bonobogenomecomparedwith}) was used as a reference
for \bonobo %
.
\end{sloppypar}

>>>>

Previous studies of some species (i.e., \human, \mouse, \cow, and
\elephant) have provided some preliminary answers to these questions.
They revealed a stunning divergence in the genetics of the placenta,
especially in later stages of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman,Hou.ea2012:Elephanttranscriptomeprovidesinsights}}.
For example, each superorder has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine}. These include
independent expansions of the prolactins in \mouse, \carmeli, \galili,
and \cow, \acp{PSG} in \human and \mouse, \acp{PAG} in \cow, genes
related to the thymus-specific serine protease \textit{PRSS16} in
\elephant (in this work), and the expansion of the growth hormone and
chorionic gonadotropin beta genes in anthropoid primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between \human and
\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}.

To address our questions, we 
sequenced and quantified the total RNA of mammalian placentas at term,
including \human, \bonobo (the sister
species to \textit{Pan troglodytes}), \smonkey,
\mouse, \galili, \carmeli, \dog, \cow, \elephant,
\armadillo, and \opossum.
These species represent four major branches of eutherian mammals
(Euarchontoglires, Laurasiatheria, Afrotheria, and Xenartha) as well
as marsupials. These placentas also represent a wide array of
placental morphologies as well as traits such as number and size of
offspring and length of gestation (\cref{tab:gestational_periods}). We
use this data, along with existing RNA-Seq data from \mouse
\cite{Wang.ea2011:surveyfornovelimprinted}, \human
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual},
\horse
\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
\cite{Archibald.ea2010:sheepgenomereferencesequence}, and
\pig \cite{Ahanda.ea2012:Predictionofaltered3-} to: 1)
identify the non-housekeeping genes which are expressed in the
placentas of all species analyzed and therefore represent the
\emph{core functionality of the placenta}; 2) identify genes whose
expression level changes significantly in each of the morphology
types.

@@

Fetal tissue was collected from the placenta of nine mammalian
species (\human [human], \bonobo [bonobo], \smonkey [spider monkey], \dog [domestic dog], \cow
[domestic cow], \elephant [African savanna elephant], \opossum [gray
short-tailed opossum], \galili [blind mole rat from Upper Galilee],
\carmeli [blind mole rat from northern Israel]
\cite{Nevo2001:adaptiveradiationofblindsubterraneanmole}). Information
on specimen ID, and collector is shown for each sample in Supplemental
\cref{tab:srr_numbers}. In the case of \bonobo, \cow, and \elephant,
the placenta was collected at term upon expulsion from the mother
following delivery. Placental tissue from \human was collected upon
delivery via a cesarean section. Near-term placental tissue from \dog
was collected upon the death of the mother and fetus from a car accident.
Placental tissue from \opossum was sampled at fetal stage 32, roughly
one day prior to birth. Only the fetal portion of the placenta was
sampled at the maternal-fetal interface (i.e. villous portion for
\human, \bonobo, \dog, and \elephant; cotyledon for \cow; diffuse
placenta for \opossum). All tissues were collected and stored in
RNAlater (Life Technologies, Carlsbad, CA) until RNA isolation could
be performed. Tissues were homogenized in Trizol, and total RNA was
isolated following the manufacturer's instructions. RNA was further
purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in conjunction
with the RNase-Free DNase Set (Qiagen, Valencia, CA), following the
manufacturer’s protocol to remove any contaminating DNA from the
sample. Total RNA was then evaluated on a Nanodrop 1000 (Thermo
Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100 (Santa
Clara, CA) to confirm concentration and quality of the isolation. RNA
quality was assessed using A260/A280 and A260/A230 absorbance values.

@@

For more complete evolutionary comparisons additional sequences of
placenta in \mouse [mouse]
(n=\Sexpr{star.logs[grepl("musculus",species),length(unique(file))]})
\cite{Wang.ea2011:surveyfornovelimprinted}, \human
(n=\Sexpr{star.logs[grepl("sapiens",species),length(unique(file))]})
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} \horse [horse]
\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep [sheep] 
\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \pig [pig]
\cite{Ahanda.ea2012:Predictionofaltered3-} were obtained from the
\ac{SRA} (Supplemental \cref{tab:srr_numbers}) and analyzed using the
same methodology as RNAs sequenced specifically for this study.

@@

\begin{sloppypar}
Reads quality was evaluated using FASTQC v0.11.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Transcriptomes without a suitable reference genome (\smonkey, \galili,
\carmeli; italic reference in Supplemental
\cref{tab:ref_genome_aligned_species}) were assembled using Trinity
v2014-07-17 \cite{Haas.ea2013:Denovotranscriptsequence} and annotated
using Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against
either the \human (for \smonkey) or \mouse (for both Spalacines) transcriptome (Supplemental
\cref{tab:ref_genome_aligned_species}), and the
the match with the highest bitscore was chosen as the annotation.
Transcripts where the highest scoring match had a bitscore below 40
were left un-annotated.
% 
All transcriptomes were aligned to the reference (or
assembled) genome using STAR v2.4.2a~%
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} (Supplemental
\cref{fig:transcriptome_statistics,fig:transcriptome_reads_vs_read_length})
and expression values in \ac{FPKM}
% \cite{Trapnell.ea2010:Transcriptassemblyandquantification}
were
quantified using Cufflinks v2.2.1
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions} (Supplemental
\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}).
%
The genome of the closely-related \textit{P. troglodytes} (separation
$\approx 1$ million years
\cite{Prüfer.ea2012:bonobogenomecomparedwith}) was used as a reference
for \bonobo %
.
\end{sloppypar}
